---
title: "Modelling tree-fungi interactions with a LBM"
author: S. Robin
date: feb 2022
output: html_document
# output: pdf_document
# geometry: margin=1.5cm
# classoption: a4paper
---

# Data

* $n = 145$ fungal species, $p = 51$ tree species.

```{r treeSpecies, echo=FALSE}
load('FungiTreeInteractions.Rdata')
head(fungiSpecies); n <- nrow(fungiSpecies)
head(treeSpecies); p <- nrow(treeSpecies)
```

* Interaction variable: $Y_{ij} = 1$  ($1 \leq i \leq n, 1 \leq j \leq p$) if fungal $i$ is aparasite of tree $j$, and 0 otherwise.

```{r data, echo=FALSE}
image(1:n, 1:p, fungiTreeInteractions, xlab='fungi', ylab='tree', main='Interaction matrix')
```

# Latent blockmodel (LBM)

Assuming $K, L$ clusters, 
\begin{align*}
  \{U_i\}_{1 \leq i \leq n} & \text{ iid} & 
  U_i & \sim \mathcal{M}{(1, \pi)} \\
  \{V_i\}_{1 \leq j \leq p} & \text{ iid} & 
  V_j & \sim \mathcal{M}{(1, \rho)} \\
  \{Y_{ij}\}_{1 \leq i, j \leq n} & \text{ indep.} \mid U, V & 
  (y_{ij} \mid U_i=k, V_j = \ell) & \sim \mathcal{B}{(\gamma_{k\ell})} \\
\end{align*}

##### Choosing the number of clusters 

```{r lbmSelect, echo=FALSE}
library(blockmodels)
# LBM <- BM_bernoulli(membership_type='LBM', adj=fungiTreeInteractions, plotting='', verbosity=0)
# LBM$estimate()
# save(LBM, file='FungiTreeLBM.Rdata')
load('FungiTreeLBM.Rdata')
plot(LBM$ICL, type='b', pch=20, xlab='K + L', ylab='ICL')
KL <- which.max(LBM$ICL)
abline(v=KL, col=2, lwd=2)
```

##### Parameter estimates

```{r lbmParms, echo=FALSE}
tau <- LBM$memberships[[KL]]$Z1; K <- ncol(tau)
eta <- LBM$memberships[[KL]]$Z2; L <- ncol(eta)
cat('K =', K, ', L =', L, '\n')
cat('pi = \n')
print(colMeans(tau))
cat('rho = \n')
print(colMeans(eta))
cat('gamma = \n')
print(LBM$model_parameters[[KL]]$pi)
```

##### Content of the clusters

```{r lbmClassif, echo=FALSE}
rankTau <- tau%*%(1:K); fungiClusterLBM <- apply(tau, 1, which.max); nk <- table(fungiClusterLBM)
rankEta <- eta%*%(1:L); treeClusterLBM <- apply(eta, 1, which.max); pl <- table(treeClusterLBM)
image(1:n, 1:p, fungiTreeInteractions[order(rankTau), order(rankEta)], 
      xlab='fungi', ylab='tree', main='Interaction matrix')
abline(v=.5+cumsum(nk)[-K], h=.5+cumsum(pl)[-L], col=4)
```
